---
title: "Getting Started with UtilsGGSV"
output:
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Getting Started with UtilsGGSV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

UtilsGGSV provides utility functions for plotting in R using `ggplot2`. This vignette demonstrates the main functions of the package.

```{r setup}
library(UtilsGGSV)
library(ggplot2)
library(magrittr)
theme_set(cowplot::theme_cowplot())
```

## Correlation Plots with `ggcorr`
 
The `ggcorr` function creates scatterplots with correlation coefficients overlaid. This is the primary function of the package and is useful for visualizing relationships between groups of measurements.

### Basic Usage

```{r ggcorr-basic}
set.seed(3)
response_vec_a <- rnorm(5)
response_tbl <- data.frame(
  group = rep(letters[1:3], each = 5),
  response = c(
    response_vec_a,
    response_vec_a * 1.2 + rnorm(5, sd = 0.2),
    response_vec_a * 2 + rnorm(5, sd = 2)
  ),
  pid = rep(paste0("id_", 1:5), 3)
)

ggcorr(
  data = response_tbl %>% dplyr::filter(group %in% c("a", "b")),
  grp = "group",
  y = "response",
  id = "pid"
)
```

### Multiple Correlation Methods

You can display multiple correlation coefficients simultaneously:

```{r ggcorr-multiple}
ggcorr(
  data = response_tbl %>% dplyr::filter(group %in% c("a", "b")),
  grp = "group",
  y = "response",
  id = "pid",
  corr_method = c("spearman", "pearson")
)
```

### Comparing Multiple Groups

When comparing more than two groups, the function creates pairwise comparisons:

```{r ggcorr-groups}
ggcorr(
  data = response_tbl,
  grp = "group",
  y = "response",
  id = "pid",
  corr_method = "kendall"
)
```

### Concordance Correlation Coefficient

The concordance correlation coefficient is useful when assessing agreement between two methods:

```{r ggcorr-concordance}
ggcorr(
  data = response_tbl %>% dplyr::filter(group %in% c("a", "b")),
  grp = "group",
  y = "response",
  id = "pid",
  corr_method = "concordance",
  abline = TRUE,
  limits_equal = TRUE
)
```

### Customizing Appearance

Text placement, font size, and other visual elements can be customized:
 
```{r ggcorr-custom}
ggcorr(
  data = response_tbl %>% dplyr::filter(group %in% c("a", "b")),
  grp = "group",
  y = "response",
  id = "pid",
  corr_method = c("spearman", "pearson", "concordance"),
  abline = TRUE,
  limits_equal = TRUE,
  coord = c(0.4, 0.17),
  font_size = 3,
  skip = 0.04,
  pval_signif = 2,
  est_signif = 2,
  ci_signif = 2
)
```

### Axis Transformations

The text placement remains consistent even when axes are transformed:

```{r ggcorr-trans}
ggcorr(
  data = response_tbl %>% dplyr::mutate(response = abs(response + 1)^4),
  grp = "group",
  y = "response",
  id = "pid",
  corr_method = "spearman",
  abline = TRUE,
  limits_equal = TRUE,
  trans = "log10",
  skip = 0.06
)
```

## Managing Axis Limits with `axis_limits`

The `axis_limits` function helps manage axis limits, particularly useful for forcing equal limits on both axes or expanding axis coordinates.

```{r axis-limits}
data("cars", package = "datasets")

p0 <- ggplot(cars, aes(speed, dist)) +
  cowplot::background_grid(major = "xy") +
  geom_point() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Axes unadjusted") +
  labs(x = "Speed", y = "Distance")

p1 <- axis_limits(
  p = p0,
  limits_equal = TRUE
) +
  labs(title = "Axes limits equal")

p2 <- axis_limits(
  p = p0,
  limits_expand = list(
    x = c(0, 50),
    y = c(-10, 200)
  )
) +
  labs(title = "Axes limits expanded")

cowplot::plot_grid(p0, p1, p2)
```

## Adding Text Annotations with `add_text_column`

The `add_text_column` function adds text annotations to plots at consistent positions, regardless of underlying axis transformations.

```{r add-text-column}
data_mod <- data.frame(x = rnorm(mean = 1, 10)^2)
data_mod$y <- data_mod$x * 3 + rnorm(10, sd = 0.5)
fit <- lm(y ~ x, data = data_mod)
coef_tbl <- coefficients(summary(fit))
results_vec <- c(
  paste0(
    "Intercept: ",
    signif(coef_tbl[1, "Estimate"][[1]], 2),
    " (",
    signif(coef_tbl[1, 1][[1]] - 2 * coef_tbl[1, 2][[1]], 3),
    ", ",
    signif(coef_tbl[1, 1][[1]] + 2 * coef_tbl[1, 2][[1]], 3),
    "; p = ",
    signif(coef_tbl[1, 4][[1]], 3),
    ")"
  ),
  paste0(
    "Slope: ",
    signif(coef_tbl[2, "Estimate"][[1]], 2),
    " (",
    signif(coef_tbl[2, 1][[1]] - 2 * coef_tbl[2, 2][[1]], 3),
    ", ",
    signif(coef_tbl[2, 1][[1]] + 2 * coef_tbl[2, 2][[1]], 3),
    "; p = ",
    signif(coef_tbl[2, 4][[1]], 3),
    ")"
  )
)

p <- ggplot(
  data = data_mod,
  aes(x = x, y = y)
) +
  geom_point() +
  cowplot::background_grid(major = "xy")

add_text_column(
  p = p,
  x = data_mod$x,
  y = data_mod$y,
  text = results_vec,
  coord = c(0.05, 0.95),
  skip = 0.07
)
```

## Cluster-Specific Plots

The `plot_cluster_*` family of functions helps visualise the characteristics of
clusters identified by an unsupervised learning method. The examples below use
the `palmerpenguins` dataset: we first apply k-means clustering to
bill and flipper measurements, then explore the clusters using each function.

```{r cluster-setup, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
set.seed(42)
penguins <- palmerpenguins::penguins
penguins <- penguins[complete.cases(penguins[
  c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g")
]), ]
vars_penguin <- c(
  "bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"
)
km <- kmeans(scale(penguins[vars_penguin]), centers = 3)
penguins$cluster <- paste0("K", km$cluster)
```

### Heat Maps with `plot_cluster_heatmap`

The `plot_cluster_heatmap` function creates a heat map where each tile shows
the percentile of the median value of a variable for a cluster. This percentile
is compared against the empirical cumulative distribution function (ECDF) of
that variable across all observations not in the cluster. Clusters and
variables are ordered along the axes via hierarchical clustering.

```{r plot-cluster-heatmap, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_heatmap(penguins, cluster = "cluster", vars = vars_penguin)
```

### Automatic Colour Palettes

The `plot_cluster_density`, `plot_cluster_scatter`, and `plot_cluster_mst`
functions (and their `plot_group_*` equivalents) automatically choose a group
colour palette via the `palette_cluster` argument (default `"auto"`):

| Groups | Palette | Notes |
|--------|---------|-------|
| 1 – 8  | Okabe-Ito | Colorblind-safe |
| 9 – 12 | ColorBrewer Paired | Light/dark pairs of 6 hues |
| 13 – 21 | Kelly (Polychrome) | Maximum-contrast sequence |
| 22 – 31 | Glasbey (Polychrome) | Algorithmically spaced colours |
| > 31   | `hue_pal()` | Fallback with a warning |

The `Polychrome` package is optional; if it is not installed the Kelly/Glasbey
tiers fall back to `hue_pal()` with a warning. You can override the automatic
selection for any group count by setting `palette_cluster` explicitly:

```{r palette-override, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
# Force the Paired palette for the 3-cluster solution
plot_cluster_density(
  penguins,
  cluster  = "cluster",
  vars     = vars_penguin,
  n_col    = 2,
  palette_cluster = "paired"
)
```

### Density Plots with `plot_cluster_density`

The `plot_cluster_density` function visualises, for each variable, how each
cluster's observations are distributed relative to the overall population. The
`density` argument controls what is shown:

- `"both"` (default): overall density plus per-cluster density curves.
- `"overall"`: the overall density with per-cluster median lines.
- `"cluster"`: one density curve per cluster.

#### Overall and per-cluster density curves (default)

```{r plot-cluster-density-default, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_density(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  n_col = 2
)
```

#### Overall density with cluster median lines

```{r plot-cluster-density-overall, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_density(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  density = "overall",
  n_col = 2
)
```

#### Per-cluster density curves only

```{r plot-cluster-density-cluster, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_density(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  density = "cluster",
  n_col = 2
)
```

#### Scaling cluster density curves

When `density` is `"both"` or `"cluster"`, the `scale` argument controls how
cluster curves are scaled relative to the overall density:

- `"max_overall"` (default): each cluster density is rescaled so that its
  maximum equals the maximum of the overall density. Y-axis values reflect the
  overall density.
- `"max_cluster"`: no rescaling; the y-axis is determined by the tallest curve.
- `"free"`: no rescaling (equivalent to `"max_cluster"`).

```{r plot-cluster-density-both-free, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
# scale = "max_cluster": natural scale for all curves
plot_cluster_density(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  density = "both",
  scale = "max_cluster",
  n_col = 2
)
```

### Scatter Plot with `plot_cluster_scatter`

The `plot_cluster_scatter` function creates a biaxial scatter plot with
observations coloured by cluster and median centroids overlaid. It supports
four dimensionality reduction methods via the `dim_red` argument: `"none"`,
`"pca"`, `"tsne"`, and `"umap"`.

#### PCA projection (default for more than two variables)

```{r plot-cluster-scatter-pca, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_scatter(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  dim_red = "pca"
)
```

#### Raw variables (`dim_red = "none"`)

When exactly two variables are needed directly as axes:

```{r plot-cluster-scatter-none, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_scatter(
  penguins,
  cluster = "cluster",
  vars = c("bill_length_mm", "flipper_length_mm"),
  dim_red = "none"
)
```

#### t-SNE

t-SNE is available when the `Rtsne` package is installed:

```{r plot-cluster-scatter-tsne, eval = requireNamespace("palmerpenguins", quietly = TRUE) && requireNamespace("Rtsne", quietly = TRUE)}
plot_cluster_scatter(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  dim_red = "tsne"
)
```

#### UMAP

UMAP is available when the `umap` package is installed:

```{r plot-cluster-scatter-umap, eval = requireNamespace("palmerpenguins", quietly = TRUE) && requireNamespace("umap", quietly = TRUE)}
plot_cluster_scatter(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  dim_red = "umap"
)
```

### Minimum-Spanning Tree with `plot_cluster_mst`

The `plot_cluster_mst` function computes the minimum-spanning tree (MST) over
clusters, using Euclidean distance between cluster median profiles. Clusters
are positioned in two dimensions via classical multidimensional scaling (MDS).
For each variable, a separate plot shows each cluster node filled by the
ECDF-standardised percentile of that cluster's median — the same colour scale
used by `plot_cluster_heatmap`. By default a named list of plots is returned;
supplying `n_col` or `n_row` returns a combined `cowplot::plot_grid` figure
with variable names as labels.

```{r plot-cluster-mst, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
plot_cluster_mst(
  penguins,
  cluster = "cluster",
  vars = vars_penguin,
  n_col = 2
)
```

## Transformations with `get_trans`

The `get_trans` function returns transformation objects from the `scales` package. It adds support for additional transformations like higher roots and `asinh`:

```{r get-trans}
get_trans("log10")
get_trans("asinh")
get_trans("root_cube")
```

